Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-07 manuscript/plos/review_1.tex
fixed many reviewers' issues

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
with the GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* hector.climente@riken.jp

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. An elegant explanation has been proposed with the omnigenic model: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. Our pipeline can easily be applied to other diseases.

\linenumbers

@@

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

Network-based biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In our studied population, only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Fig~\ref{fig:benchmark}B), while providing more interpretable solutions.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}}
  \begin{tabular}{lrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \(\rho\)\textsubscript{consensus} }\\
  \thickhline
  HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
  \hline
  dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
  heinz & 4 & 3 & 113\,633 & 0.001 & 0.21\\
  HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
  LEAN & 0 & 0 & - & - & -\\
  SConES GI & 0 (1) & 0 & - & - & -\\
  SigMod & 142 & 249 & 92\,603 & 0.008 & 0.73\\
  \hline
  Consensus & 93 & 186 & 50\,737 & 0.006 & 1\\
  Stable consensus & 68 & 49 & 94\,854 & 0.005 & 0.54 \\
  \end{tabular}
  \begin{flushleft} \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P\=/value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

\begin{figure}[!ht]
  \centering
  \includegraphics[height=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant network (BH adjusted P\=/value~<~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

Despite their differences, there are additional common themes. All obtained solution subnetworks have lower association P\=/values than the whole PPIN (median P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solution subnetworks presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value < 0.03). Second, the genes in four solution subnetworks displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value < 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (Supplementary table~\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

\subsection{A case study: the consensus network}
\label{results:consensus}

Despite the heterogeneity of the solutions, their shared properties suggest that each method captures different aspects of cancer susceptibility. Indeed, only 20 genes are common to more than two solutions (\nameref{sfig:consensus_stats}A), but encouragingly, the more methods selected a gene, the higher its association score to the phenotype (\nameref{sfig:consensus_stats}B). To leverage on their strengths and compensate their respective weaknesses, we built a consensus subnetwork that captures the mechanisms most shared among the solution subnetworks (Section~\ref{methods:methods}). This subnetwork (Fig~\ref{fig:consensus}) contains 93 genes and exhibits the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.7\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus subnetwork on GENESIS (Section~\ref{methods:methods}) }. Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in \nameref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

>>>>

\usepackage[shortcuts]{extdash}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Biological networks and GWAS: comparing and combining network methods to understand the genetics of familial breast cancer susceptibility in the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
with the GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* hector.climente(at)riken.jp

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. Indeed, it allows to jointly consider the statistical association between genetic variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new breast cancer susceptibility genes by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution, their predictive power, and the stability and the runtime of the methods. Network methods include prior knowledge in the analysis: they might tolerate selecting a gene with a high P-value, as long as it connects other two, low P-value genes. Thanks to this, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, at least two network methods selected 14 neighbors of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To account for that, we produce a stable consensus solution, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than the average gene. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an elegant explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}: they tend to occupy central positions in the network, interconecting different modules TODO. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each gene has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints, and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods on GWAS data. They do so by using different interpretations of the omnigenic model, providing a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). Two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16} the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In GENESIS only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

Closer examination reveals two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all the GENESIS genotypes as input, and the phenotype (cancer/healthy) as outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:snp2gene}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model is low (sensitivity = 55\%, specificity = 55\%, Section \ref{methods:benchmark}). Together, these issues motivate exploring network methods, which consider not only statistical association, but also the location of each gene in a PPIN to find susceptibility genes.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value $<$ 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}A). Despite that, the methods tend to agree on the genes with the strongest signal: genes selected by more methods tended to have lower P-value of association (Fig~\ref{fig:solution_overview}B). Another prominent difference is the solution size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI did not recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution network to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the PPIN. The first row contains the summary statistics on the whole PPIN.}}
  \begin{tabular}{lrrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \# components } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \# genes in consensus } \\
  \thickhline
  HINT HT           & 13\,619 & 142\,541  & 15  & 16\,706   & 0.46  & 93/93 \\
  \hline
  dmGWAS            & 194     & 450       & 1   & 49\,115   & 0.19  & 55/93 \\
  heinz             & 4       & 3         & 1   & 113\,633  & 0.001 & 4/93  \\
  HotNet2           & 440     & 374       & 130 & 7\,739    & 0.048 & 63/93 \\
  LEAN              & 0       & 0         & 0   & -         & -     & 0/93  \\
  SConES GI         & 0 (1)   & 0         & 0   & -         & -     & 0/93  \\
  SigMod            & 142     & 249       & 11  & 92\,603   & 0.008 & 84/93 \\
  \hline
  Consensus         & 93      & 186       & 21  & 50\,737   & 0.006 & 93/93 \\
  Stable consensus  & 68      & 49        & 32  & 94\,854   & 0.005 & 43/93 \\
  \end{tabular}
  \begin{flushleft} \textbf{\# genes}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. {\bf \# components }: number of connected components. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median VEGAS2 P\=/value of the selected genes. \textbf{\# genes in consensus}: Number of genes in common between the method's solution, and the 93 genes in the consensus solution.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant network (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(B)}~VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(D)}~Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

Despite their differences, there are additional common themes. All obtained solutions have lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This exemplifies the trade-off between controlling for type I error and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solutions presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in four solutions displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value$<$1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

\subsection{A case study: the consensus solution}
\label{results:consensus}

Despite their shared properties, the differences between the solutions of the different methods suggest that each of them captures different aspects of cancer susceptibility. Indeed, out of the 668 genes that are selected by at least one method, only 93 are selected by at least two, 20 by three, and none by four or more. Encouragingly, the more methods selected a gene, the higher its association score to the phenotype (Fig \ref{fig:solution_overview}B), a relationship that plateaus at 2. Hence, to leverage on their strengths and compensate their respective weaknesses, we built a consensus solution that captures the genes shared among at least two solutions (Section~\ref{methods:methods}). This solution (Fig~\ref{fig:consensus}) contains 93 genes and exhibits the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.7\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). The latter ones are also colored in pink. All gene names are indicated in \nameref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}
